A memory frontier for complex synapses: Supplementary material

### 2013-11-06 NIPS/SynapseNIPSsupp.tex
typo in nips supp
final NIPS paper

Although the theory and derivations outlined in the main text are relatively self-contained, here we provide more details for the aid of the reviewer.

@@

In this section we'll provide a summary of all the relevant properties of ergodic Markov chains in continuous time to define notation and .
It is a straightforward generalisation of material that can be found in \cite{kemeny1960finite} with some ideas from \cite{hunter2000survey}.

@@

A Markov process is described by a matrix of transitions rates, $\MM_{ij}$, from state $i$ to $j$.
The probabilities of being in each state at time $t$, the row-vector $\mathbf{p}(t)$, evolve according to
%
\begin{equation}\label{eq:rowsum}
  \diff{\mathbf{p}(t)}{t} = \mathbf{p}(t) \MM,
  \qquad
  \mathbf{p}(t)\onev=1,
  \qquad
  \MM\onev=0.
\end{equation}
%

@@

@@

Suppose that the Markov process, defined by $\MM$, depends on some parameter $\alpha$.
Differentiating \eqref{eq:funddef} gives
%
\begin{equation}\label{eq:diffZ}
  \diff{\fund}{\alpha} = \fund \diff{\MM}{\alpha} \fund.
\end{equation}
%
We can substitute this into the derivative of \eqref{eq:fundprob}:
%
\begin{equation}\label{eq:diffp}
  \diff{\eq}{\alpha} = \pib \fund \diff{\MM}{\alpha} \fund = \eq \diff{\MM}{\alpha} \fund.
\end{equation}
%
We can rewrite this in component form and use the fact that $\MM_{ii} = - \sum_{i\neq j} \MM_{ij}$:
%
\begin{equation}\label{eq:diffpT}
\begin{aligned}
  \diff{\eq_k}{\alpha} &= \sum_{i,j} \eq_i \diff{\MM_{ij}}{\alpha} \fund_{jk} \\
    &= \sum_{i\neq j} \eq_i \diff{\MM_{ij}}{\alpha} \fund_{jk} + \sum_i \eq_i \diff{\MM_{ii}}{\alpha} \fund_{ik} \\
    &= \sum_{i\neq j} \eq_i \diff{\MM_{ij}}{\alpha} (\fund_{jk} - \fund_{ik}) \\
    &= \sum_{i\neq j} \diff{\MM_{ij}}{\alpha} \eq_i \eq_k (\fptb_{ik} - \fptb_{jk}),
\end{aligned}
\end{equation}
%
which is the result of \cite{cho2000markov} that we need.
Note that the summand vanishes for $i=j$, so we can drop the restriction $i\neq j$ from the range of the sum.

@@

This allows us to write the lumpability condition \eqref{eq:lump}, and the transition matrix for the lumped process compactly:
%
\begin{equation}\label{eq:lumpcompact}
  VU\MM V = \MM V,
  \qquad
  \widehat{\MM} = U\MM V.
\end{equation}
%
By induction, one can show that similar relations hold for all powers:
%
\begin{equation}\label{eq:lumppower}
  VU\MM^nV = \MM^nV,
  \qquad
  \widehat{\MM}^n = U\MM^n V,
\end{equation}
%
and, via the Taylor series, for the exponential as well:
%
\begin{equation}\label{eq:lumpexp}
  VU \e^{t\MM} V = \e^{t\MM}V,
  \qquad
  \e^{t\widehat{\MM}} = U\e^{t\MM} V.
\end{equation}
%
The equilibrium distribution of the lumped process is given by
%
\begin{equation}\label{eq:lumpeq}
  \widehat{\mathbf{p}}^\infty = \eq V.
\end{equation}
%
\section{Signal-to-Noise ratio (SNR)}\label{sec:SNR}

In this section we will look at the signal-to-noise curve, and put an upper bound on its initial value.
We will only consider ergodic Markov chains.
Transient states would be unoccupied in equilibrium and would not be accessed by the signal creation process, therefore they could be removed from the analysis.
Absorbing chains are degenerate cases: they have zero initial signal but infinite decay times, so they can only be approached as the limit of a sequence of ergodic chains.

@@

Which results in our final formula:
%
\begin{equation}\label{eq:SNRcurve}
  \SNR(t) = \sqrt{N} (2f\pot f\dep )\, \eq (\W\pot  - \W\dep )\,\e^{rt\frg}\w.
\end{equation}
%
The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.

@@

In this section we will find an upper bound on the area under the signal-to-noise curve.
As in \sref{sec:SNR}, we will only consider ergodic Markov chains.
We will see in \sref{sec:multistate} that the optimal chain is absorbing, so it lies on the boundary of the (open) set of ergodic chains, but it still puts an upper bound on the area.

@@

It is also helpful to define the following quantities:
%
\begin{equation}\label{eq:areacoeffs}
  \begin{aligned}
    c_k &= \diff{\ln\eq_k}{\alpha}
      = \sum_{ij} \eq_i \enc_{ij} \prn{\fptb_{ik}-\fptb_{jk}}
      = - \prn{\eq q \fptb}_k
      = \frac{(\eq q \fund)_k}{\eq_k}, \\
    a_i &= \sum_j \enc_{ij} \eq_i (\eta^+_{i} - \eta^+_{j}),\\
    \implies
    \hat{A} &= \sum_k c_k \eq_k \w_k
      = 2\sum_i a_i.
  \end{aligned}
\end{equation}
%
Note that the optimal choice of $\w$ is $\w_k = \sgn(c_k)$.

@@

@@

@@

@@

@@

If we put the states in order of decreasing $\eta^+_k$, the results of the \sref{sec:areacoeffincr} tell us that the $c_k$ are non-decreasing.
This implies that the last term of the final expression in \eqref{eq:shortcutarea} is non-negative.
If it is non-zero (there would need to be a lot of degeneracy for it to be zero), this would imply that $\mu\potdep _{g,g\pm m}>0$, which in turn implies that $\W\potdep _{g,g\pm m}=0$.
This would tell us that the process with the maximal area has to have a multi-state topology.

@@

The multi-state/serial topology is defined by (see \cite{amit1994learning,Fusi2007multistate,Leibold2008serial}):
%
\begin{equation}\label{eq:multistatedef}
  \W\pot _{ij} = q\pot _i \delta_{i+1,j},
  \qquad
  \W\dep _{ij} = q\dep _j \delta_{i,j+1}.
\end{equation}
%
It saturates various inequalities:
%
\begin{equation}\label{eq:multistateineq}
  \begin{aligned}
    \fptb_{ik} - \fptb_{jk} &=
      \begin{cases}
        \fptb_{ij},  &\text{if}\quad i \leq j \leq k \quad\text{or}\quad i \geq j \geq k,\\
        -\fptb_{ji}, &\text{if}\quad j \leq i \leq k \quad\text{or}\quad j \geq i \geq k,\\
      \end{cases} \\
    r\eq_i \frg_{ij} \prn{\fptb_{ij}+\fptb_{ji}} &= 1 \quad\text{if}\quad i=j\pm1,
  \end{aligned}
\end{equation}
%
and it satisfies detailed balance (a.k.a.\ reversibility a.k.a. $\CL^2_{\eq}$ self-adjointness):
%
\begin{equation}\label{eq:multistateprob}
  f\pot  q\pot _i \eq_i = f\dep  q\dep _i \eq_{i+1},
\end{equation}
%
which means we can always choose the transition rates, $q\potdep _i$, to give any desired equilibrium probabilities, $\eq_i$.

This allows us to calculate the $c_k$'s:
%
\begin{equation}\label{eq:areacoeffchain}
\begin{aligned}
  c_k =&\, \sum_{i<k} \fpt_{i,i+1} \prn{\eq_{i}q\pot _i+\eq_{i+1}q\dep _i}
    %\\&
    - \sum_{i\geq k} \fpt_{i+1,i} \prn{\eq_{i}q\pot _i+\eq_{i+1}q\dep _i}
  ,\\
  c_{k+1} - c_k =&\, \prn{\fpt_{k,k+1}+\fpt_{k+1,k}} \prn{\frac{\eq_{k}\frg_{k,k+1}}{f\pot }+\frac{\eq_{k+1}\frg_{k+1,k}}{f\dep }}
    =%\\=&\,
    \frac{1}{rf\pot f\dep },\\
  \sum_k c_k \eq_k =&\, \sum_{ij} \eq_i \enc_{ij} (\eta-\eta) = 0,\\
%  \implies c_k =&\, \frac{\prn{k-\frac{n+1}{2}} - \sum_j\prn{j-\frac{n+1}{2}}\eq_j}{f\pot f\dep },
  \implies c_k =&\, \frac{k - \sum_j j\eq_j}{rf\pot f\dep },
\end{aligned}
\end{equation}
%
where we used \eqref{eq:multistateineq} to derive the first two equations respectively and Th.\ref{th:kemenyconst} to derive the third. This allows us to write the area as
%
\begin{equation}\label{eq:multistatearea}
 % A = 2\sqrt{N} \sum_k \brk{\prn{k-\frac{n+1}{2}} - \sum_j\prn{j-\frac{n+1}{2}}\eq_j} \eq_k \w_k.
  A = \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we used $\w_k=\sgn(c_k)$, as discussed after \eqref{eq:areacoeffs}.
%It will also help to define $\eq_\pm = \sum_k \eq_k \prn{\frac{1\pm\w_k}{2}}$.
%Note that
%%
%\begin{equation}\label{eq:allpplusorminus}
%  \begin{aligned}
%    \eq_+ &=1
%     \quad&\implies\quad
%     \eq_k \w_k &= \eq_k
%     \quad&\implies\quad
%     A=0,\\
%    \eq_- &=1
%     \quad&\implies\quad
%     \eq_k \w_k &= -\eq_k
%     \quad&\implies\quad
%     A=0,
%  \end{aligned}
%\end{equation}
%%
%neither of which are optimal.
%
%Consider the Lagrangian
%%
%\begin{equation}\label{eq:multistatelagrangian}
%  \CL = \frac{A}{2\sqrt{N}} + \lambda\prn{1-\sum_i \eq_i} + \sum_i \mu_i \eq_i,
%\end{equation}
%%
%where $\lambda$ is a Lagrange multiplier and $\mu_i$ are Kuhn-Tucker multipliers (see \eqref{eq:extremum}). Extremising \wrt $\eq_i$:
%%
%\begin{equation}\label{eq:multistateext}
%%  \pdiff{\CL}{\eq_i} = \brk{\prn{i-\frac{n+1}{2}} - \sum_j\prn{j-\frac{n+1}{2}}\eq_j} \w_i
%%    - \prn{i-\frac{n+1}{2}}\prn{\eq_+-\eq_-} - \lambda + \mu_i = 0.
%  \pdiff{\CL}{\eq_i} = \brk{i - \sum_j j\eq_j} \w_i
%    - i\prn{\eq_+-\eq_-} - \lambda + \mu_i = 0.
%\end{equation}
%%
%Consider an $i$ such that $\w_i=\w_{i+1}$:
%%
%\begin{equation}\label{eq:multistatemudiff}
%  \mu_{i+1} - \mu_i = \prn{\eq_+-\eq_-} - \w_i.
%\end{equation}
%%
%As $\abs{\eq_+-\eq_-}<1$ (unless we want $A=0$, \eqref{eq:allpplusorminus}), this means the $\mu_i$ increase amongst the negative $\w_i$ and decrease amongst the positive $\w_i$.
%Therefore the only non-zero equilibrium probabilities at optimum are $\eq_1$ and $\eq_n$.

First let us maximise \eqref{eq:multistatearea} at fixed $\eq_\pm = \sum_k \eq_k \prn{\frac{1\pm\w_k}{2}}$.
Clearly this will happen when we put all of the probability at the ends: $\eq_1=\eq_-$ and $\eq_n=\eq_+$ are the only non-zero $\eq_k$.
This gives an area of
%\footnote{Note that including the dropped factors in \eqref{eq:noisepm} would only change this to $A \leq \sqrt{N(4\eq_+\eq_-)}(n-1)/r$, which would have no effect on what follows.}

%
\begin{equation}\label{eq:multistateextarea}
  A \leq \frac{\sqrt{N}}{r}(M-1)\prn{4\eq_+\eq_-}.
\end{equation}
%
This is maximised at $\eq_+=\eq_-=\half$:
%
\begin{equation}\label{eq:maxarea}
  A \leq \frac{\sqrt{N}}{r}(M-1).
\end{equation}
%
Note that the chain that achieves this is not ergodic, the two states at each end are absorbing. This is similar to the results found numerically in \cite{Barrett2008discrete} in a slightly different situation.

>>>>

Here we provide more details underlying the derivations of results in the main paper.

@@

In this section we'll provide a summary of all the relevant properties of ergodic Markov chains in continuous time to define notation.
It is a generalization of material that can be found in \cite{kemeny1960finite} with some ideas from \cite{hunter2000survey}.

@@

A continuous time Markov process is described by a matrix of transitions rates, $\MM_{ij}$, from state $i$ to $j$ with row sums equal to zero ($\MM\onev=0$).
The probabilities of being in each state at time $t$, the row-vector $\mathbf{p}(t)$, evolve according to
%
\begin{equation}\label{eq:rowsum}
  \diff{\mathbf{p}(t)}{t} = \mathbf{p}(t) \MM,
\end{equation}
%
where $\mathbf{p}(t)\onev=1$.

@@

For our results below regarding the integral of the memory curve, it can be useful to invert the stochastic transition matrix, $\MM$.
However, since $\MM$ has a zero eigenvalue, it cannot be inverted.
For this reason, the fundamental matrix arises as a useful surrogate for the inverse of $\MM$.
It is related to the first passage times, as we will see in the \hyperref[sec:fpt]{next subsection}.
Here we define the fundamental matrix and review its properties.

@@

Suppose that the Markov process, defined by $\MM$, depends on some parameter $\alpha$.
Differentiating \eqref{eq:funddef} gives
%
\begin{equation}\label{eq:diffZ}
  \diff{\fund}{\alpha} = \fund \diff{\MM}{\alpha} \fund.
\end{equation}
%
We can substitute this into the derivative of \eqref{eq:fundprob}:
%
\begin{equation}\label{eq:diffp}
  \diff{\eq}{\alpha} = \pib \fund \diff{\MM}{\alpha} \fund = \eq \diff{\MM}{\alpha} \fund.
\end{equation}
%
We can rewrite this in component form and use the fact that $\MM_{ii} = - \sum_{i\neq j} \MM_{ij}$:
%
\begin{equation}\label{eq:diffpT}
\begin{aligned}
  \diff{\eq_k}{\alpha} &= \sum_{i,j} \eq_i \diff{\MM_{ij}}{\alpha} \fund_{jk} \\
    &= \sum_{i\neq j} \eq_i \diff{\MM_{ij}}{\alpha} \fund_{jk} + \sum_i \eq_i \diff{\MM_{ii}}{\alpha} \fund_{ik} \\
    &= \sum_{i\neq j} \eq_i \diff{\MM_{ij}}{\alpha} (\fund_{jk} - \fund_{ik}) \\
    &= \sum_{i\neq j} \diff{\MM_{ij}}{\alpha} \eq_i \eq_k (\fptb_{ik} - \fptb_{jk}).
\end{aligned}
\end{equation}
%
This is a generalization of a result of \cite{cho2000markov} from discrete to continuous time that we will need below.
Note that the summand vanishes for $i=j$, so we can drop the restriction $i\neq j$ from the range of the sum.

@@

This allows us to write the lumpability condition \eqref{eq:lump}, and the transition matrix for the lumped process compactly:
%
\begin{equation}\label{eq:lumpcompact}
  VU\MM V = \MM V,
  \qquad
  \widehat{\MM} = U\MM V.
\end{equation}
%
By induction, one can show that similar relations hold for all powers:
%
\begin{equation}\label{eq:lumppower}
  VU\MM^nV = \MM^nV,
  \qquad
  \widehat{\MM}^n = U\MM^n V,
\end{equation}
%
and, via the Taylor series, for the exponential as well:
%
\begin{equation}\label{eq:lumpexp}
  VU \e^{t\MM} V = \e^{t\MM}V,
  \qquad
  \e^{t\widehat{\MM}} = U\e^{t\MM} V.
\end{equation}
%
The equilibrium distribution of the lumped process is given by
%
\begin{equation}\label{eq:lumpeq}
  \widehat{\mathbf{p}}^\infty = \eq V.
\end{equation}
%

\section{Signal-to-Noise ratio (SNR)}\label{sec:SNR}

In this section we will look at the signal-to-noise curve, and put an upper bound on its initial value.
We need only consider ergodic Markov chains.
Transient states would be unoccupied in equilibrium and would not be accessed by the signal creation process, therefore they could be removed from the analysis.
Absorbing chains are degenerate cases: they have zero initial signal but infinite decay times, so they can only be approached as the limit of a sequence of ergodic chains.

@@

This results in our final formula:
%
\begin{equation}\label{eq:SNRcurve}
  \SNR(t) = \sqrt{N} (2f\pot f\dep )\, \eq (\W\pot  - \W\dep )\,\e^{rt\frg}\w.
\end{equation}
%
The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.

@@

In this section we will find an upper bound on the area under the signal-to-noise curve.
As in \sref{sec:SNR}, we will only consider ergodic Markov chains.
We will see in \sref{sec:multistate} that the optimal chain is absorbing, so it lies on the boundary of the (open) set of ergodic chains, but it still puts an upper bound on the area for any chain in the interior.

@@

It is also helpful to define the following quantities:
%
\begin{equation}\label{eq:areacoeffs}
  \begin{aligned}
    c_k &= \diff{\ln\eq_k}{\alpha}
      = \sum_{ij} \eq_i \enc_{ij} \prn{\fptb_{ik}-\fptb_{jk}}
      = - \prn{\eq \enc \fptb}_k
      = \frac{(\eq \enc \fund)_k}{\eq_k}, \\
    a_i &= \sum_j \enc_{ij} \eq_i (\eta^+_{i} - \eta^+_{j}),\\
    \implies
    \hat{A} &= \sum_k c_k \eq_k \w_k
      = 2\sum_i a_i.
  \end{aligned}
\end{equation}
%
Note that the optimal choice of $\w$ is $\w_k = \sgn(c_k)$.

@@

In the following, we will mathematically define the classes of perturbations pictorially described in Figure 3 of the main paper.
In order to do so, we will need to consider expressions for derivatives of various quantities with respect to $\W\potdep_{ij}$.

@@

Here we describe the perturbations corresponding to Figure 3(a,b) of the main paper.

@@

In this subsection we will define perturbations corresponding to Figure 3(c) of the main text.

In the \hyperref[sec:areacoeffincr]{section below}, we will show that the $c_k$ are non-decreasing, if we put the states in order of decreasing $\eta^+_k$.
This implies that the last term of the final expression in \eqref{eq:shortcutarea} is non-negative.
If it is non-zero (there would need to be a lot of degeneracy for it to be zero), this would imply that $\mu\potdep _{g,g\pm m}>0$, which in turn implies that $\W\potdep _{g,g\pm m}=0$.
This would tell us that the process with the maximal area has to have a multi-state topology.

@@

In the \hyperref[sec:shortcuts]{previous subsection} we defined perturbations corresponding to Figure 3(c) of the main text.
In order to show that those perturbations increase the area, we must now show that the $c_k$ are non-decreasing, if we put the states in order of decreasing $\eta^+_k$.

@@

@@

The previous results indicate that the area under the memory curve of any model is bounded by the area under the memory curve of a model with the serial/multistate topology having the same equilibrium distribution.
Here we compute this area, which we will see depends only on this equilibrium distribution.

The multi-state/serial topology is defined by (see \cite{amit1994learning,Fusi2007multistate,Leibold2008serial}):
%
\begin{equation}\label{eq:multistatedef}
  \W\pot _{ij} = q\pot _i \delta_{i+1,j},
  \qquad
  \W\dep _{ij} = q\dep _j \delta_{i,j+1}.
\end{equation}
%
Because it has no shortcuts, it saturates various inequalities:
%
\begin{equation}\label{eq:multistateineq}
  \begin{aligned}
    \fptb_{ik} - \fptb_{jk} &=
      \begin{cases}
        \fptb_{ij},  &\text{if}\quad i \leq j \leq k \quad\text{or}\quad i \geq j \geq k,\\
        -\fptb_{ji}, &\text{if}\quad j \leq i \leq k \quad\text{or}\quad j \geq i \geq k,\\
      \end{cases} \\
    r\eq_i \frg_{ij} \prn{\fptb_{ij}+\fptb_{ji}} &= 1 \quad\text{if}\quad i=j\pm1,
  \end{aligned}
\end{equation}
%
and it satisfies detailed balance (a.k.a.\ reversibility a.k.a. $\CL^2_{\eq}$ self-adjointness):
%
\begin{equation}\label{eq:multistateprob}
  f\pot  q\pot _i \eq_i = f\dep  q\dep _i \eq_{i+1},
\end{equation}
%
which means we can always choose the transition rates, $q\potdep _i$, to give any desired equilibrium probabilities, $\eq_i$.

This allows us to calculate the $c_k$'s:
%
\begin{equation}\label{eq:areacoeffchain}
\begin{aligned}
  c_k =&\, \sum_{i<k} \fpt_{i,i+1} \prn{\eq_{i}q\pot _i+\eq_{i+1}q\dep _i}
    %\\&
    - \sum_{i\geq k} \fpt_{i+1,i} \prn{\eq_{i}q\pot _i+\eq_{i+1}q\dep _i}
  ,\\
  c_{k+1} - c_k =&\, \prn{\fpt_{k,k+1}+\fpt_{k+1,k}} \prn{\frac{\eq_{k}\frg_{k,k+1}}{f\pot }+\frac{\eq_{k+1}\frg_{k+1,k}}{f\dep }}
    =%\\=&\,
    \frac{1}{rf\pot f\dep },\\
  \sum_k c_k \eq_k =&\, \sum_{ij} \eq_i \enc_{ij} (\eta-\eta) = 0,\\
%  \implies c_k =&\, \frac{\prn{k-\frac{n+1}{2}} - \sum_j\prn{j-\frac{n+1}{2}}\eq_j}{f\pot f\dep },
  \implies c_k =&\, \frac{k - \sum_j j\eq_j}{rf\pot f\dep },
\end{aligned}
\end{equation}
%
where we used \eqref{eq:multistateineq} to derive the first two equations respectively and Th.\ref{th:kemenyconst} to derive the third. This allows us to write the area as
%
\begin{equation}\label{eq:multistatearea}
 % A = 2\sqrt{N} \sum_k \brk{\prn{k-\frac{n+1}{2}} - \sum_j\prn{j-\frac{n+1}{2}}\eq_j} \eq_k \w_k.
  A = \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we used $\w_k=\sgn(c_k)$, as discussed after \eqref{eq:areacoeffs}.
This reproduces equation (15) of the main paper.

%It will also help to define $\eq_\pm = \sum_k \eq_k \prn{\frac{1\pm\w_k}{2}}$.
%Note that
%%
%\begin{equation}\label{eq:allpplusorminus}
%  \begin{aligned}
%    \eq_+ &=1
%     \quad&\implies\quad
%     \eq_k \w_k &= \eq_k
%     \quad&\implies\quad
%     A=0,\\
%    \eq_- &=1
%     \quad&\implies\quad
%     \eq_k \w_k &= -\eq_k
%     \quad&\implies\quad
%     A=0,
%  \end{aligned}
%\end{equation}
%%
%neither of which are optimal.
%
%Consider the Lagrangian
%%
%\begin{equation}\label{eq:multistatelagrangian}
%  \CL = \frac{A}{2\sqrt{N}} + \lambda\prn{1-\sum_i \eq_i} + \sum_i \mu_i \eq_i,
%\end{equation}
%%
%where $\lambda$ is a Lagrange multiplier and $\mu_i$ are Kuhn-Tucker multipliers (see \eqref{eq:extremum}). Extremising \wrt $\eq_i$:
%%
%\begin{equation}\label{eq:multistateext}
%%  \pdiff{\CL}{\eq_i} = \brk{\prn{i-\frac{n+1}{2}} - \sum_j\prn{j-\frac{n+1}{2}}\eq_j} \w_i
%%    - \prn{i-\frac{n+1}{2}}\prn{\eq_+-\eq_-} - \lambda + \mu_i = 0.
%  \pdiff{\CL}{\eq_i} = \brk{i - \sum_j j\eq_j} \w_i
%    - i\prn{\eq_+-\eq_-} - \lambda + \mu_i = 0.
%\end{equation}
%%
%Consider an $i$ such that $\w_i=\w_{i+1}$:
%%
%\begin{equation}\label{eq:multistatemudiff}
%  \mu_{i+1} - \mu_i = \prn{\eq_+-\eq_-} - \w_i.
%\end{equation}
%%
%As $\abs{\eq_+-\eq_-}<1$ (unless we want $A=0$, \eqref{eq:allpplusorminus}), this means the $\mu_i$ increase amongst the negative $\w_i$ and decrease amongst the positive $\w_i$.
%Therefore the only non-zero equilibrium probabilities at optimum are $\eq_1$ and $\eq_n$.

In order to obtain an upper bound on the area under the memory curve of any model, we now maximise the area of the serial model with respect to its equilibrium distribution.
First let us maximise \eqref{eq:multistatearea} at fixed $\eq_\pm = \sum_k \eq_k \prn{\frac{1\pm\w_k}{2}}$.
Clearly this will happen when we put all of the probability at the ends: $\eq_1=\eq_-$ and $\eq_n=\eq_+$ are the only non-zero $\eq_k$.
This gives an area of
%\footnote{Note that including the dropped factors in \eqref{eq:noisepm} would only change this to $A \leq \sqrt{N(4\eq_+\eq_-)}(n-1)/r$, which would have no effect on what follows.}

%
\begin{equation}\label{eq:multistateextarea}
  A \leq \frac{\sqrt{N}}{r}(M-1)\prn{4\eq_+\eq_-}.
\end{equation}
%
This is maximised at $\eq_+=\eq_-=\half$:
%
\begin{equation}\label{eq:maxarea}
  A \leq \frac{\sqrt{N}}{r}(M-1).
\end{equation}
%
This yields the area bound of equation (16) of the main text.

Note that the chain that achieves this is not ergodic, the two states at each end are absorbing. This is similar to the results found numerically in \cite{Barrett2008discrete} in a slightly different situation.
